---
title: "Graunt, Halley, and US 1993 Life Table : Summary"
author: "coop711"
date: '`r Sys.Date()`'
output: html_document
editor_options: 
  chunk_output_type: console
---

# Source of Data

<img src = "../pics/graunt_table.png" width = "240"/>

<!--## Data Input-->

<P style = "page-break-before:always">

# R base graphics

## Graunt

<!--
```{r, graunt data input}
graunt <- data.frame(x = c(0, seq(6, 76, by = 10)), 
                     xPo_g = c(100, 64, 40, 25, 16, 10, 6, 3, 1))
```

### More data

* US 1993 life table for the same age group

```{r, us93 data input}
us93 <- data.frame(x = graunt$x, 
                   xPo_us = c(100, 99, 99, 98, 97, 95, 92, 84, 70))
```


```{r one data frmae}
(graunt_us <- data.frame(graunt, 
                         xPo_us = us93$xPo))
```

### Life Expectancy

The basic principle is that the area under the survival function is the life expectancy. 

$X \ge 0$, $X \sim F(x)$ => $X \equiv F^{-1}(U), U \sim U(0,1)$, therefore, 

$E(X) = E\{F^{-1}(U)\} = \int_{0}^{1} F^{-1}(u)du = \int_0^{\infty} 1-F(x) dx = \int_{0}^{\infty} S(x) dx$


### Setting up coordinates for `polygon()` (Clockwise)

```{r, polygon coordinates}
(graunt_x <- c(graunt$x, 0))
(graunt_y <- c(graunt$xPo_g, 0))
graunt_poly <- data.frame(x = graunt_x, y = graunt_y)
```
-->

```{r, title axis labels, fig.width = 6, fig.height = 6, echo = FALSE}
png(filename = "../pics/graunt_plots.png", width = 640, height = 960)
par(mfrow = c(3, 2))
plot(graunt, ann = FALSE, xaxt = "n", yaxt = "n", type = "b")
plot(graunt, ann = FALSE, xaxt = "n", yaxt = "n", type = "b")
axis(side = 1, at = graunt$x, labels = graunt$x)
axis(side = 2, at = graunt$xPo_g, labels = graunt$xPo_g)
plot(graunt, ann = FALSE, xaxt = "n", yaxt = "n", type = "b")
axis(side = 1, at = graunt$x, labels = graunt$x)
axis(side = 2, at = graunt$xPo_g, labels = graunt$xPo_g)
abline(v = c(0, 76), lty = 2)
plot(graunt, ann = FALSE, xaxt = "n", yaxt = "n", type = "b")
axis(side = 1, at = graunt$x, labels = graunt$x)
axis(side = 2, at = graunt$xPo_g, labels = graunt$xPo_g)
abline(v = c(0, 76), lty = 2)
polygon(graunt_poly, density = 15)
plot(graunt, ann = FALSE, xaxt = "n", yaxt = "n", type = "b")
axis(side = 1, at = graunt$x, labels = graunt$x)
axis(side = 2, at = graunt$xPo_g, labels = graunt$xPo_g)
abline(v = c(0, 76), lty = 2)
polygon(graunt_poly, density = 15)
abline(v = graunt$x, lty = 2)
plot(graunt, ann = FALSE, xaxt = "n", yaxt = "n", type = "b")
axis(side = 1, at = graunt$x, labels = graunt$x)
axis(side = 2, at = graunt$xPo_g, labels = graunt$xPo_g)
abline(v = c(0, 76), lty = 2)
polygon(graunt_poly, density = 15)
abline(v = graunt$x, lty = 2)
points(graunt, pch = 21, col = "black", bg = "white")
# dev.copy(png, file = "../pics/graunt_plots.png", width = 720, height = 1080)
dev.off()
plot(graunt, ann = FALSE, xaxt = "n", yaxt = "n", type = "b")
axis(side = 1, at = graunt$x, labels = graunt$x)
axis(side = 2, at = graunt$xPo_g, labels = graunt$xPo_g)
abline(v = c(0, 76), lty = 2)
polygon(graunt_poly, density = 15)
abline(v = graunt$x, lty = 2)
points(graunt, pch = 21, col = "black", bg = "white")
main_title <- "Graunt's Survival Function"
x_lab <- "Age (years)"
y_lab <- "Proportion of Survival (%)"
title(main = main_title, xlab = x_lab, ylab = y_lab)
dev.copy(png, file = "../pics/graunt_surv.png", 480, 480)
dev.off()
```

<!--### Area under the curve-->

<!--
*  `diff()`, `head()`, and `tail()` can be used to write a function to compute the area easily.  

```{r, area function}
area.R <- function(x, y) {
  sum(diff(x) * (head(y, -1) + tail(y, -1))/2)
  }
```
-->

The area under the curve can be approximated by the sum of the areas of trapezoids, therefore the area is $\sum_{i=1}^{n-1} (x_{i+1}-x_i)\times\frac{1}{2}(y_i + y_{i+1})$. Therefore, the life expectancy of Graunt's life table is 
`r area.R(graunt$x, graunt$xPo_g)/100`(years).

<P style = "page-break-before:always">

## Graunt and US 1993

The shaded area between the survival function of Graunt and that of US 1993 represents the difference of life expectancies.

<!--

### Setting coordinates for `polygon()`

```{r, polygon coordinates us 93 added}
us_graunt_x <- c(us93$x, rev(graunt$x))
us_graunt_y <- c(us93$xPo_us, rev(graunt$xPo_g))
us_graunt <- data.frame(x = us_graunt_x, y = us_graunt_y)
```
-->

```{r, title-axis-labels, fig.width = 6, fig.height = 6, echo = FALSE}
plot(graunt, ann = FALSE, xaxt = "n", yaxt = "n", type = "b")
axis(side = 1, at = graunt$x, labels = graunt$x)
axis(side = 2, at = graunt$xPo, labels = graunt$xPo_g)
abline(v = c(0, 76), lty = 2)
lines(us93, type = "b")
abline(h = 70, lty = 2)
axis(side = 2, at = 70, labels = 70, las = 1)
polygon(us_graunt, density = 15, col = "blue", border = NA)
abline(v = graunt$x, lty = 2)
points(us_graunt, pch = 21, col = "black", bg = "white")
main_title_g_us <- "Survival Function of Graunt and US 1993"
title(main = main_title_g_us, xlab = x_lab, ylab = y_lab)
```

The area under the US 1993 survival function is `r area.R(us93$x, us93$xPo_us)/100`, so, the area of shaded region, that is the difference of life expectancy, is `r area.R(us93$x, us93$xPo_us)/100 - area.R(graunt$x, graunt$xPo_g)/100` (years).

<P style = "page-break-before:always">

## Part of Halley's life table

```{r, Halley lifetable, echo = FALSE}
age <- 0:84
lx <- c(1238, 1000, 855, 798, 760, 732, 710, 692, 680, 670, 661, 653, 646, 640, 634, 628, 622, 616, 610, 604, 598, 592, 586, 579, 573, 567, 560, 553, 546, 539, 531, 523, 515, 507, 499, 490, 481, 472, 463, 454, 445, 436, 427, 417, 407, 397, 387, 377, 367, 357, 346, 335, 324, 313, 302, 292, 282, 272, 262, 252, 242, 232, 222, 212, 202, 192, 182, 172, 162, 152, 142, 131, 120, 109, 98, 88, 78, 68, 58, 50, 41, 34, 28, 23, 20)
halley <- data.frame(age, lx)
halley$xPo <- round(halley$lx / lx[1] * 100, digits = 1)
```

```{r, Halley lifetable with xPo, echo = FALSE}
library(knitr)
kable(cbind(head(halley, n = 10), 
            tail(halley, n = 10)), 
      align = c("c", "r", "r", "c", "r", "r"), 
      caption = "Halley's Lifetable")
```

```{r, Halley with xPo only, echo = FALSE, results = 'hide'}
halley_lx <- halley[-3]
halley <- halley[-2]
head(halley)
tail(halley)
```

<P style = "page-break-before:always">

## Graunt and Halley

<!--


```{r, data subsetting}
age_graunt <- age %in% graunt$x
halley_graunt <- subset(halley, age_graunt)
xPo_halley_age_6 <- halley$xPo[age == 6]
main_title_2 <- "Survival Function of Graunt and Halley"
```

## Polygon

Setting the coordinates for `polygon()`. The intersection is found at `x = 10.8, y = 52.8` with `locator(1)` and couple of trial and errors.

* Upper region

```{r, coords upper region}
poly_1_x <- c(graunt$x[1:2], 10.8, halley$age[11:1])
poly_1_y <- c(graunt$xPo_g[1:2], 52.8, halley$xPo[11:1])
poly_upper <- data.frame(x = poly_1_x, y = poly_1_y)
```

* Lower region

```{r, coords lower region}
poly_2_x <- c(10.8, halley$age[12:85], graunt$x[9:3])
poly_2_y <- c(52.8, halley$xPo[12:85], graunt$xPo_g[9:3])
poly_lower <- data.frame(x = poly_2_x, y = poly_2_y)
```

--> 

```{r, fill the points, fig.width = 6, fig.height = 6, echo = FALSE}
plot(halley, ann = FALSE, xaxt = "n", yaxt = "n", type = "l")
abline(v = c(0, 76, 84), lty = 2)
points(halley_graunt, pch = 21, col = "black", bg = "white")
lines(graunt, type = "b", pch = 21, col = "black", bg = "white")
axis(side = 1, at = c(graunt$x, 84), labels = c(graunt$x, 84))
axis(side = 2, at = graunt$xPo_g, labels = graunt$xPo_g, las = 1)
axis(side = 2, at = xPo_halley_age_6, labels = xPo_halley_age_6, las = 1)
text(x = c(16, 36), y = c(20, 50), label = c("Graunt", "Halley"))
title(main = main_title_2, xlab = x_lab, ylab = y_lab)
polygon(poly_upper, angle = 45, density = 15, col = "red", border = NA)
polygon(poly_lower, angle = 45, density = 15, col = "green", border = NA)
points(graunt, pch = 21, col = "black", bg = "white")
points(halley_graunt, pch = 21, col = "black", bg = "white")
points(x = 84, y = halley$xPo[85], pch = 21, col = "black", bg = "white")
```

<!--### Life expectancy-->

Compute the difference of life expectancies, Halley's is `r area.R(halley$age, halley$xPo)/100` (years), and Graunt's is `r area.R(graunt$x, graunt$xPo_g)/100` (years).

<P style = "page-break-before:always">

## Graunt, Halley, and US 1993

<!--Polygon with R Base Plot-->

<!--
### Coordinates 

In order to make the graphs truncated at the age 76, restrict the age of Halley up to 76.

```{r, polygon data}
graunt_2 <- graunt
halley_2 <- halley
us93_2 <- us93
names(graunt_2) <- c("x", "Graunt")
names(halley_2) <- c("x", "Halley")
names(us93_2) <- c("x", "US93")
poly_lower_76 <- subset(poly_lower, poly_lower$x <= 76)
poly_3_x <- c(us93_2$x, halley_2$x[85:12], 10.8, graunt_2$x[2:1])
poly_3_y <- c(us93_2$US93, halley_2$Halley[85:12], 52.8, graunt_2$Graunt[2:1])
poly_us <- data.frame(x = poly_3_x, y = poly_3_y)
poly_us_76 <- subset(poly_us, poly_us$x <= 76)
```

### Straight to Polygon
-->

```{r, Graunt Halley and US 1993 together, fig.width = 6, fig.height = 6, echo = FALSE}
plot(halley, ann = FALSE, xaxt = "n", yaxt = "n", type = "l")
abline(v = c(0, 76, 84), lty = 2)
points(halley_graunt, pch = 21, col = "black", bg = "white")
lines(graunt, type = "b", pch = 21, col = "black", bg = "white")
lines(us93, type = "b", pch = 21, col = "black", bg = "white")
axis(side = 1, at = c(graunt$x, 84), labels = c(graunt$x, 84))
axis(side = 2, at = c(graunt$xPo_g, xPo_halley_age_6), labels = c(graunt$xPo_g, xPo_halley_age_6), las = 1)
abline(h = 70, lty = 2)
axis(side = 2, at = 70, labels = 70, las = 1)
main_title_3 <- "Survival Function Plots"
title(main = main_title_3, xlab = x_lab, ylab = y_lab)
polygon(poly_upper, angle = 45, density = 15, col = "red", border = NA)
polygon(poly_lower_76, angle = 45, density = 15, col = "green", border = NA)
polygon(poly_us_76, angle = 45, density = 15, col = "blue", border = NA)
points(graunt, pch = 21, col = "black", bg = "white")
points(halley_graunt, pch = 21, col = "black", bg = "white")
points(us93_2, pch = 21, col = "black", bg = "white")
points(x = 84, y = halley$xPo[85], pch = 21, col = "black", bg = "white")
text(x = c(16, 36, 70), y = c(25, 50, 90), label = c("Graunt", "Halley", "US93"))
```

<P style = "page-break-before:always">

# ggplot 

<!--
```{r, library ggplot2}
library(ggplot2)
```

## Data Reshape

Attach `reshape2` package to change wide format to long format

```{r, reshape2}
library(reshape2)
```

How `melt()` works

```{r, format of melt}
graunt_us_melt <- melt(graunt_us, 
                       id.vars = "x", 
                       measure.vars = c("xPo_g", "xPo_us"), 
                       value.name = "xPo", variable.name = "times")
graunt_us_melt
str(graunt_us_melt)
```

* Change factor levels of `times`

```{r, factor level}
levels(graunt_us_melt$times) <- c("Graunt", "US1993")
str(graunt_us_melt)
```
-->

## Graunt

<!--
## Structure of ggplot

```{r, geom aes, fig.width = 6, fig.height = 6, echo = FALSE}
g1 <- ggplot() + 
  geom_line(data = graunt, 
            mapping = aes(x = graunt$x, y = graunt$xPo))
g2 <- g1 +
  geom_point(data = graunt, 
             mapping = aes(x = graunt$x, y = graunt$xPo), 
             shape = 21, fill = "white")
g3 <- g2 +
  theme_bw()
g4 <- g3 +
   xlab(x_lab) + 
   ylab(y_lab) + 
   ggtitle(main_title) +
   scale_x_continuous(breaks = graunt$x) + 
   scale_y_continuous(breaks = graunt$xPo_g)
g5 <- g4 +
  geom_vline(xintercept = graunt$x, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted")
pg4 <- g4 +
  geom_polygon(data = graunt_poly, 
               mapping = aes(x = x, y = y), 
               alpha = 0.7, fill = "grey")
```
-->

```{r, graunt polygon ggplot, fig.width = 6, fig.height = 6, echo = FALSE}
pg4
```

<P style = "page-break-before:always">

## Graunt and US 1993

<!--### Points and Lines

Step by step approach to understand the grammar of ggplot

* We set `ggplot()` to accept varying `data.frame()` and `aes()`in `geom_polygon`

```{r, geom_point geom_line, fig.width = 6, fig.height = 6}
gu1 <- ggplot() + 
  geom_line(data = graunt_us_melt, 
            mapping = aes(x = x, y = xPo, colour = times))
gu2 <- gu1 + 
  geom_point(data = graunt_us_melt, 
             mapping = aes(x = x, y = xPo, colour = times), 
             shape = 21, fill = "white")
gu3 <- gu2 + 
  theme_bw()
```

## Polygon 

Reuse `us_graunt` which contains `x = us_graunt_x` and `y = us_graunt_y` for `polygon()`. Note that we start with `gu3`, and also note how to remove default legends.

```{r, polygon, fig.width = 6, fig.height = 6}
gup3 <- gu3 + 
  geom_polygon(data = us_graunt, 
               mapping = aes(x = x, y = y), 
               alpha = 0.3, fill = "blue")
gup4 <- gup3 + 
  guides(colour = "none")
```

## Change default annotations

### Points and Lines

1. Change the x-axis and y-axis labels
    
```{r, axis labels, fig.width = 6, fig.height = 6}
gu4 <- gu3 + 
   xlab(x_lab) + 
   ylab(y_lab)
```

2. Add main title

```{r, ggtitle, fig.width = 6, fig.height = 6}
gu4 <- gu3 + 
   xlab(x_lab) + 
   ylab(y_lab) + 
   ggtitle(main_title_g_us)
```

3. Change legend title
    
```{r, legends title, fig.width = 6, fig.height = 6}
gu4 <- gu3 + 
   xlab(x_lab) + 
   ylab(y_lab) + 
   ggtitle(main_title_g_us) +
   labs(colour = "Era")
```

4. Change legends.
    
```{r, change legends, fig.width = 6, fig.height = 6}
gu4 <- gu3 + 
   xlab(x_lab) + 
   ylab(y_lab) +
   ggtitle(main_title_g_us) +
   labs(colour = "Era") +
   scale_colour_discrete(labels = c("Graunt Era", "US 1993"))
```

5. Place legends inside the plot

```{r, legends inside, fig.width = 6, fig.height = 6}
gu5 <- gu4 + 
   theme(legend.position = c(0.8, 0.5))
```

6. Change x-axis and y-axis tick marks

```{r, axis tick marks, fig.width = 6, fig.height = 6}
gu6 <- gu5 + 
   scale_x_continuous(breaks = graunt$x) + 
   scale_y_continuous(breaks = graunt$xPo_g)
```

## Polygon

Add information to the plot drawn with `polygon()`

1. Start with `gup4`

```{r, start with p4, fig.width = 6, fig.height = 6}
gup4
```

2. Main title, x-axis and y-axis labels

```{r, title and axis labels,  fig.width = 6, fig.height = 6}
gup5 <- gup4 + 
   xlab(x_lab) + 
   ylab(y_lab) +
   ggtitle(main_title_g_us)
```

3. `"Graunt Era"`, `"US 1993"`, `"Difference of Life Expectancies"` at proper positions
    
```{r, text polygon,  fig.width = 6, fig.height = 6}
gup6 <- gup5 + 
   annotate("text", 
            x = c(20, 40, 70), y = c(20, 60, 90), 
            label = c("Graunt Era", "Difference of\nLife Expectancies", "US 1993"))
```

4. x-axis and y-axis tick marks-->

```{r, familiar axis labels, fig.width = 6, fig.height = 6, echo = FALSE}
(gup7 <- gup6 + 
   scale_x_continuous(breaks = graunt$x) + 
   scale_y_continuous(breaks = graunt$xPo_g))
```

<P style = "page-break-before:always">

## Graunt and Halley

<!--
### Data Reshaping

Since the observed ages are different, we need final structure of the data frame to be melted. So, create copies of `graunt` and `halley` and extract parts of what we need and give feasible names.  

```{r, data reshaping}
graunt_halley_melt <- melt(list(graunt_2, halley_2), 
                           id.vars = "x", 
                           value.name = "xPo", 
                           variable.name = "Who")
str(graunt_halley_melt)
graunt_halley_melt <- graunt_halley_melt[-4]
(graunt_halley_melt_g <- subset(graunt_halley_melt, 
                                graunt_halley_melt$x %in% graunt$x))
```
 
## Survival Function, Step by Step

```{r, first plot for Graunt and Halley, fig.width = 6, fig.height = 6}
(gh1 <- ggplot() + 
  geom_line(data = graunt_halley_melt, 
            mapping = aes(x = x, y = xPo, colour = Who)))
(gh2 <- gh1 + 
  geom_point(data = graunt_halley_melt_g, 
             mapping = aes(x = x, y = xPo, colour = Who), 
             shape = 21, fill = "white"))
(gh3 <- gh2 + 
  theme_bw() + 
  xlab(x_lab) + 
  ylab(y_lab) + 
  ggtitle(main_title_2))
(gh4 <- gh3 + 
  theme(legend.position = c(0.8, 0.8)) +
  annotate("text", 
           x = c(16, 36), y = c(20, 50), 
           label = c("Graunt", "Halley")) +
  scale_x_continuous(breaks = c(graunt$x, 84)) + 
  scale_y_continuous(breaks = c(graunt$xPo_g, xPo_halley_age_6)))
```

## Polygon

Reuse `poly_upper` data frame and `poly_lower` data frame.

```{r, polygon for Graunt and Halley, fig.width = 6, fig.height = 6}
(ghp4 <- gh4 + 
  geom_polygon(data = poly_upper, 
               mapping = aes(x = x, y = y), 
               alpha = 0.3, fill = "red"))
(ghp5 <- ghp4 + 
  geom_polygon(data = poly_lower, 
               mapping = aes(x = x, y = y), 
               alpha = 0.3, fill = "green"))
```
-->

```{r, ghp, fig.width = 6, fig.height = 6, echo = FALSE}
(ghp5 <- ghp5 +
  geom_point(data = data.frame(x = 84, y = halley$xPo[85]), 
             mapping = aes(x = x, y = y),  
             colour = 3, shape = 21, fill = "white"))
```

<P style = "page-break-before:always">

## Graunt, Halley, and US93

<!--
## Data Reshape

```{r, reshape stage 3}
# us93_2 <- us93
# names(us93_2) <- c("x", "US93")
ghu_melt <- melt(list(graunt_2, halley_2, us93_2), 
                 id.vars = "x", 
                 value.name = "xPo", 
                 variable.name = "Who")
ghu_melt_g <- ghu_melt[ghu_melt$x %in% graunt$x, ]
# main_title_3 <- "Survival Function Plots"
```

## Survival Function Plots with ggplot

```{r, Survival function plot for Graunt Halley and US93, fig.width = 6, fig.height = 6}
ghu1 <- ggplot() + 
  geom_line(data = ghu_melt, 
            mapping = aes(x = x, y = xPo, colour = Who))
ghu2 <- ghu1 + 
  geom_point(data = ghu_melt_g, 
             mapping = aes(x = x, y = xPo, colour = Who), 
             shape = 21, fill = "white")
ghu3 <- ghu2 + 
  theme_bw() + 
  xlab(x_lab) + 
  ylab(y_lab) + 
  ggtitle(main_title_3)
ghu4 <- ghu3 + 
  theme(legend.position = c(0.2, 0.2)) +
  annotate("text", 
           x = c(36, 36, 70), y = c(25, 50, 90), 
           label = c("Graunt", "Halley", "US93")) +
  scale_x_continuous(breaks = c(graunt$x, 84)) + 
  scale_y_continuous(breaks = c(graunt$xPo_g, xPo_halley_age_6))
```

## Polygon

```{r, polygon for Graunt Halley and US93, fig.width = 6, fig.height = 6}
ghup4 <- ghu4 + 
  geom_polygon(data = poly_upper, 
               mapping = aes(x = x, y = y), 
               alpha = 0.3, fill = "red")
ghup5 <- ghup4 + 
  geom_polygon(data = poly_lower_76, 
               mapping = aes(x = x, y = y), 
               alpha = 0.3, fill = "green")
ghup6 <- ghup5 +
  geom_polygon(data = poly_us_76, 
               mapping = aes(x = x, y = y), 
               alpha = 0.3, fill = "blue")
```
-->

```{r, ghup, fig.width = 6, fig.height = 6, echo = FALSE}
(ghup7 <- ghup6 +
  geom_point(data = data.frame(x = 84, y = halley$xPo[85]), 
             mapping = aes(x = x, y = y),  
             colour = 3, shape = 21, fill = "white"))
```

<!--
### `dump()` and `source()`

* Check out how to save and retrieve. Use `source()` and `load()` for retrieval.

```{r, dump and save}
dump("area.R", file = "area.R")
save.image("graunt_halley_1801.RData")
```
-->